---
title: "p8105_hw6_zlb2008"
author: "Zakari L. Billo"
date: "2025-12-03"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(broom)
library(purrr)
library(p8105.datasets)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 7,
  fig.align = "center",
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## Problem 1 

Create a city_state variable (e.g. “Baltimore, MD”), and a binary variable indicating whether the homicide is solved. Omit cities Dallas, TX; Phoenix, AZ; and Kansas City, MO – these don’t report victim race. Also omit Tulsa, AL – this is a data entry mistake. For this problem, limit your analysis those for whom victim_race is white or black. Be sure that victim_age is numeric.

For the city of Baltimore, MD, use the glm function to fit a logistic regression with resolved vs unresolved as the outcome and victim age, sex and race as predictors. Save the output of glm as an R object; apply the broom::tidy to this object; and obtain the estimate and confidence interval of the adjusted odds ratio for solving homicides comparing male victims to female victims keeping all other variables fixed.

Now run glm for each of the cities in your dataset, and extract the adjusted odds ratio (and CI) for solving homicides comparing male victims to female victims. Do this within a “tidy” pipeline, making use of purrr::map, list columns, and unnest as necessary to create a dataframe with estimated ORs and CIs for each city.

Create a plot that shows the estimated ORs and CIs for each city. Organize cities according to estimated OR, and comment on the plot.

```{r}
homicide_df <- read.csv('data/homicide-data.csv', na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  mutate(
    city_state = str_c(city, ", ", state)
  ) |>
  filter(
    !city_state %in% c("Dallas, TX", "Phoenix, AZ", "Kansas City, MO", "Tulsa, AL"),
    victim_race %in% c("White", "Black")
  ) |>
  mutate(
    victim_age = as.numeric(victim_age),
    case_outcome = ifelse(disposition == "Closed by arrest", 1, 0)
  ) |>
  filter(!is.na(victim_age)) |>
  select(city_state, case_outcome, victim_age, victim_sex, victim_race)
```

glm for Baltimore, MD

```{r}
bmore_results =
  homicide_df |> 
  filter(city_state == "Baltimore, MD") |> 
  glm(
    case_outcome ~ victim_age + victim_sex + victim_race,
    family = binomial(),
    data = _
  ) |> 
  broom::tidy(conf.int = TRUE) |> 
  mutate(
    OR = exp(estimate),
    CL_low = exp(conf.low),
    CL_high = exp(conf.high)
  ) |>
  filter(term == "victim_sexMale") |>
  select(term, log_OR = estimate, OR, CL_low, CL_high, p.value)

bmore_results |> knitr::kable(digits = 3)
```
The odds ratio for solved homocides among male victims compared to that among female victims in Baltimore, MD is 0.426 (95% CI: 0.324, 0.558), adjusting for victim age and race.

```{r}
city_results =
  homicide_df |> 
  nest(data = -city_state) |> 
  mutate(
    fit = map(
      data,
      ~ glm(case_outcome ~ victim_age + victim_sex + victim_race,
            family = binomial(), data = .x)
    ),
    tidy_fit = map(fit, ~ broom::tidy(.x, conf.int = TRUE))
  ) |>
  unnest(tidy_fit) |>
  filter(term == "victim_sexMale") |>
  mutate(
    OR = exp(estimate),
    CL_low = exp(conf.low),
    CL_high = exp(conf.high)
  ) |>
  select(city_state, OR, CL_low, CL_high, p.value)

city_results |> 
  knitr::kable(digits = 3)
```
plot the ORs 

```{r}
city_results |>
  mutate(city_state = fct_reorder(city_state, OR)) |>
  ggplot(aes(y = city_state, x = OR)) +
  geom_point(size = 2) +
  geom_errorbarh(aes(xmin = CL_low, xmax = CL_high), height = 0.15) +
  geom_vline(xintercept = 1, color = "red", linetype = "dotted") +
  labs(
    title = "Adjusted Odds Ratio for Solved Homicides: Male vs Female Victims",
    x = "Odds Ratio (adjusted for age & race)",
    y = "City"
  ) +
  theme_minimal()
```

The plot of ORs by city shows that most cities have a lower odds of solving homicide cases for male victims compared to female victims. 

Albuquerque, NM; Stockton, CA; and Fresno, CA are the exception given that they have the highest estimated odds ratios, meaning that homicides with male victim are more likely to be solved than homicides with female victims. However, most of these cities' estimated ORs have confidence intervals that include the null, implying there is no statistically significant evidence of a difference in the odds of solving homicides between male and female victims.

New York, NY; Baton Rouge, LA; and Omaha, NE have the lowest odds of solving homicide cases for male victims compared to female victims. The OR confidence intervals for these cities is below 1, providing statistically significant evidence for lower solve rates among male-victim cases.

## Problem 2 


